---
title: "AB323_bioptics"
output: html_document
date: "2026-04-29"
---

## load packages
```{r}
library(tidyverse)
library(readr)
library(readxl)
library(here)
```


## import data
```{r}
#setwd("C:/Users/Eva Samson/OneDrive - University of Bergen/Dokumenter/Bergen/AB-323/bioptics")

reflectance_raw <- read_xlsx("bioptics_reflectance_v2.xlsx", skip = 11) |> 
  select(!ID)

methanol_raw <- read_xlsx("bioptics_absorbance_v2.xlsx", skip = 11, sheet = "methanol extracts") |> 
  select(!ID)

water_raw <- read_xlsx("bioptics_absorbance_v2.xlsx", skip = 11, sheet = "water extracts") |> 
  select(!ID)

invivo_raw <- read_xlsx("bioptics_absorbance_v2.xlsx", skip = 11, sheet = "in vivo") |> 
  select(!ID)

filters_raw <- read_xlsx("bioptics_absorbance_v2.xlsx", skip = 11, sheet = "filters") |> 
  select(!ID)
```


## filters
```{r}
filters_PAR <- filters_raw |> 
  filter(400 <= wavelength, wavelength <= 700) |> 
  #select(!Astand_filter) |> 
  pivot_longer(cols = !wavelength, names_to = "sample", values_to = "value")
  # |> pivot_longer(cols = starts_with("A"), names_to = "sample", values_to = "value")


```


```{r}
filters_plot <- 
  ggplot(filters_PAR, 
         aes(x = wavelength, y= value, colour = sample)) +
  scale_color_viridis_d() +
  geom_line() +
  theme_minimal()

filters_plot

#ggsave(here("filters_plot.png"), plot = filters_plot, width = 120, height = 100, units = "mm", scale = 1.2)
```


## reflectance vs absorbance

### brown algae
```{r}
refl_brownalgae <- reflectance_raw |> 
  select(wavelength, contains("RB"))
abs_brownalgae <- invivo_raw |> 
  select(wavelength, contains("AB"))

refl_abs_brownalgae <- merge(refl_brownalgae, abs_brownalgae, by = "wavelength") |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  pivot_longer(cols = starts_with("RB") | starts_with("AB"), names_to = "sample", values_to = "value") |> 
  mutate(measurement = case_when(str_detect(sample, "R") ~ "reflectance", str_detect(sample, "A") ~ "absorbance"))


refl_abs_brownalgae_plot <- refl_abs_brownalgae |> 
  ggplot(aes(x = wavelength, y = value, color = sample)) + 
  geom_line(data = subset(refl_abs_brownalgae, measurement == "reflectance")) + 
  geom_line(data = subset(refl_abs_brownalgae, measurement == "absorbance"), aes(y = value*50)) +
  scale_color_viridis_d(option = "magma") +
  scale_y_continuous(sec.axis = sec_axis(~ . /50)) +
  theme_minimal()


refl_abs_brownalgae_plot

#ggsave(here("refl_abs_brownalgae_plot.png"), plot = refl_abs_brownalgae_plot, width = 120, height = 100, units = "mm", scale = 1.2)
```



### green algae
```{r}
refl_greenalgae <- reflectance_raw |> 
  select(wavelength, contains("RG"))
abs_greenalgae <- invivo_raw |> 
  select(wavelength, contains("AG"))

refl_abs_greenalgae <- merge(refl_greenalgae, abs_greenalgae, by = "wavelength") |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  pivot_longer(cols = starts_with("RG") | starts_with("AG"), names_to = "sample", values_to = "value") |> 
  mutate(measurement = case_when(str_detect(sample, "R") ~ "reflectance", str_detect(sample, "A") ~ "absorbance"))

refl_abs_greenalgae_plot <- refl_abs_greenalgae |> 
  ggplot(aes(x = wavelength, y = value, color = sample)) + 
  geom_line(data = subset(refl_abs_greenalgae, measurement == "reflectance")) + 
  geom_line(data = subset(refl_abs_greenalgae, measurement == "absorbance"), aes(y = value*50)) +
  scale_color_viridis_d(option = "magma") +
  scale_y_continuous(sec.axis = sec_axis(~ . /50)) +
  theme_minimal()

refl_abs_greenalgae_plot
```


### red algae
```{r}
refl_redalgae <- reflectance_raw |> 
  select(wavelength, contains("RR"))
abs_redalgae <- invivo_raw |> 
  select(wavelength, contains("AR"))

refl_abs_redalgae <- merge(refl_redalgae, abs_redalgae, by = "wavelength") |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  pivot_longer(cols = starts_with("RR") | starts_with("AR"), names_to = "sample", values_to = "value") |> 
  mutate(measurement = case_when(str_detect(sample, "RR") ~ "reflectance", str_detect(sample, "AR") ~ "absorbance"))

refl_abs_redalgae_plot <- refl_abs_redalgae |> 
  ggplot(aes(x = wavelength, y = value, color = sample)) + 
  geom_line(data = subset(refl_abs_redalgae, measurement == "reflectance")) + 
  geom_line(data = subset(refl_abs_redalgae, measurement == "absorbance"), aes(y = value*50)) +
  scale_color_viridis_d(option = "magma") +
  scale_y_continuous(sec.axis = sec_axis(~ . /50)) +
  theme_minimal()

refl_abs_redalgae_plot
```

## absorbance vs. absorbance
```{r}
#brown
abs_brownalgae_ratios <- abs_brownalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = AB1/AB2, ratio2 = AB1/AB3, ratio3 = AB1/AB4, ratio4 = AB2/AB3, ratio5 = AB2/AB4, ratio6 = AB3/AB4) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
abs_brownalgae_ratios_plot <- abs_brownalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
abs_brownalgae_ratios_plot

#green
abs_greenalgae_ratios <- abs_greenalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = AG1/AG2, ratio2 = AG1/AG3, ratio3 = AG2/AG3) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
abs_greenalgae_ratios_plot <- abs_greenalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
abs_greenalgae_ratios_plot

#red
abs_redalgae_ratios <- abs_redalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = AR1/AR2, ratio2 = AR1/AR3, ratio3 = AR1/AR4, ratio4 = AR2/AR3, ratio5 = AR2/AR4, ratio6 = AR3/AR4) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
abs_redalgae_ratios_plot <- abs_redalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
abs_redalgae_ratios_plot
```


## reflectance vs. reflectance
```{r}
#brown
refl_brownalgae_ratios <- refl_brownalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = RB1/RB2, ratio2 = RB1/RB3, ratio3 = RB1/RB4, ratio4 = RB2/RB3, ratio5 = RB2/RB4, ratio6 = RB3/RB4) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
refl_brownalgae_ratios_plot <- refl_brownalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
refl_brownalgae_ratios_plot

#green
refl_greenalgae_ratios <- refl_greenalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = RG1/RG2, ratio2 = RG1/RG3, ratio3 = RG2/RG3) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
refl_greenalgae_ratios_plot <- refl_greenalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
refl_greenalgae_ratios_plot

#red
refl_redalgae_ratios <- refl_redalgae |> 
  filter(400 <= wavelength, wavelength <= 700 ) |> 
  mutate(ratio1 = RR2/RR3, ratio2 = RR2/RR4, ratio3 = RR3/RR4) |> 
  select(wavelength, starts_with("ratio")) |> 
  pivot_longer(cols = starts_with("ratio"), names_to = "ratio", values_to = "value")
refl_redalgae_ratios_plot <- refl_redalgae_ratios |> 
  ggplot(aes(x = wavelength, y = value, color = ratio)) + 
  geom_line() + 
  theme_minimal()
refl_redalgae_ratios_plot
```


## absorbance/reflectance
This displays the ratio between absorbance and reflectance. For each class of algae, a representative and well-distinguished sample was chosen for calculating the ratio between absorbance and reflectance.

### brown algae
```{r}
abs_refl_ratio_brown <- refl_abs_brownalgae |> 
  filter(sample == "RB1"| sample == "AB3") |> 
  pivot_wider(names_from = measurement, values_from = value) |> 
  select(!sample) |> 
  group_by(wavelength) |> 
  summarise(
    reflectance = na.omit(reflectance)[1],  # first non-NA in Value1
    absorbance = na.omit(absorbance)[1]   # first non-NA in Value2
  ) |> 
  mutate(ratio = absorbance/reflectance) |> 
  mutate(difference = absorbance*10 - reflectance)
  

abs_refl_ratio_brown_plot <- abs_refl_ratio_brown |> 
  ggplot(aes(x = wavelength, y = ratio)) + 
  geom_line() + 
  theme_minimal()


abs_refl_ratio_brown_plot
```

### green algae
```{r}
abs_refl_ratio_green <- refl_abs_greenalgae |> 
  filter(sample == "RG2"| sample == "AG2") |> 
  pivot_wider(names_from = measurement, values_from = value) |> 
  select(!sample) |> 
  group_by(wavelength) |> 
  summarise(
    reflectance = na.omit(reflectance)[1],  # first non-NA in Value1
    absorbance = na.omit(absorbance)[1]   # first non-NA in Value2
  ) |> 
  mutate(ratio = absorbance/reflectance) |> 
  mutate(difference = absorbance - reflectance)
  

abs_refl_ratio_green_plot <- abs_refl_ratio |> 
  ggplot(aes(x = wavelength, y = ratio)) + 
  geom_line() + 
  theme_minimal()


abs_refl_ratio_green_plot
```


### red algae
```{r}
abs_refl_ratio_red <- refl_abs_redalgae |> 
  filter(sample == "RR4"| sample == "AR1") |> 
  pivot_wider(names_from = measurement, values_from = value) |> 
  select(!sample) |> 
  group_by(wavelength) |> 
  summarise(
    reflectance = na.omit(reflectance)[1],  # first non-NA in Value1
    absorbance = na.omit(absorbance)[1]   # first non-NA in Value2
  ) |> 
  mutate(ratio = absorbance/reflectance) |> 
  mutate(difference = absorbance - reflectance)
  

abs_refl_ratio_red_plot <- abs_refl_ratio |> 
  ggplot(aes(x = wavelength, y = ratio)) + 
  geom_line() + 
  theme_minimal()


abs_refl_ratio_red_plot
```


## absorbance - reflectance
### brown algae
```{r}
abs_refl_diff_brown_plot <- abs_refl_ratio_brown |> 
  ggplot(aes(x = wavelength, y = difference)) + 
  geom_line() + 
  theme_minimal()

abs_refl_diff_brown_plot
```

### green algae
```{r}
abs_refl_diff_green_plot <- abs_refl_ratio_green |> 
  ggplot(aes(x = wavelength, y = difference)) + 
  geom_line() + 
  theme_minimal()

abs_refl_diff_green_plot
```

### red algae
```{r}
abs_refl_diff_red_plot <- abs_refl_ratio_red |> 
  ggplot(aes(x = wavelength, y = difference)) + 
  geom_line() + 
  theme_minimal()

abs_refl_diff_red_plot
```

# crap
for separating by class:
```{r}
mutate(class = case_when( str_detect(sample, "B") ~ "brown", str_detect(sample, "G") ~ "green", str_detect(sample, "AR|RR") ~ "red" )) 
```
sample modifications:
```{r}
#  filter(chl_a_ug_l_1 != "#VALUE!") |> 
#  filter(sampling_date == "24.04.2026") |> 
#  filter(filter == "GFF") |> 
#  filter(station == "Pier") |> 
#  mutate(chl_a_ug_l_1 = as.numeric(chl_a_ug_l_1)) |>
#  mutate(sample_depth_m = as.character(sample_depth_m)) |> 
#  group_by(sample_depth_m)
```

